---
title: "Data Visualization with ggplot2"
subtitle: "Statistical Computing with R"
author: "Waseem Medhat"
date: "`r Sys.Date()`"
logo: "content/r-logo.svg"
output:
  ioslides_presentation:
    wide: true
---

The colorful part of the course
-------------------------------

- `ggplot2` is an elegant package for drawing graphs.
- One of the biggest competitive advantages of R compared to other data analytic
  software.
- It has a layer-based syntax.
- Based on the Grammar of Graphics: a philosophy of looking at graphs.

Philosophy
----------

<div class="red2">
"What is a graph? And how does it relate to the data?"
</div>

- A graph is a visual representation of data, in which data are mapped to
  _aesthetic_ properties (e.g. length, color, X/Y coordinates) of some
  _geometric_ shapes (e.g. points, bars, lines).

- In a simple scatterplot:
    + The geometric shape is a point/circle.
    + The aesthetics are X/Y coordinates.

- In a bar chart:
    + The geometric shape is a bar/rectangle.
    + The aesthetics are X coordinates and bar height.

General syntax
--------------

- `ggplot()` is used to put the base layer given two arguments: data to be
  plotted (must be a data frame) and aesthetic mapping `aes()`.
- Each additional layer is added with a plus sign and a `geom_****()` function
  to add some shapes.
- Other non-geom functions can be added for customization.

```{r message=FALSE}
# always remember to load the package
library(ggplot2)
```

```
ggplot(<data_frame>, aes(x = <a_variable>, <...>)) +
  geom_<something>(<...>)
```

Scatterplots
------------

```{r fig.height=4}
# first layer         aesthetics              geometric shape
ggplot(data = mtcars, aes(x = wt, y = mpg)) + geom_point()
```

Histograms
----------

```{r fig.height = 4, message=FALSE}
ggplot(data = mtcars, aes(x = mpg)) + geom_histogram()
```

Histograms
----------

```{r fig.height = 4, message=FALSE}
ggplot(data = mtcars, aes(x = mpg)) + geom_histogram(binwidth = 5)
```

Scatterplots with colors
------------------------

```{r fig.height=4}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```

Non-aesthetics
--------------

- You can change properties without mapping to a variable by adding arguments
  _outside_ `aes()`.

```{r fig.height=3}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width)) +
  geom_point(color = "red")
```

Scatterplots with fit line
--------------------------

```{r fig.height=3}
ggplot(data = mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Scatterplots with fit line
--------------------------

```{r fig.height=3}
ggplot(data = iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

Bar charts
----------

```{r echo=FALSE, message=FALSE, include=FALSE}
knitr::knit("part_ii_basics.Rmd")
```

```{r fig.height=4}
ggplot(data = statisticians, aes(x = older_than_50, fill = disease)) +
  geom_bar()
```

Color vs. fill
--------------

- Line and point shapes have only one color. This color is mapped to the `color`
  aesthetic

- Bar shapes have two colors: one for the border, which is mapped to the `color`
  aesthetic, and another for the bar itself, which is mapped to `fill`.

Boxplots
--------

```{r fig.height=4}
ggplot(statisticians, aes(x = disease, y = age)) + geom_boxplot()
```


Facets
------

- There is a `ggplot2` function that splits a plot into panels (or facets)
  according to a grouping variable.

Facets
------

```{r fig.height=3}
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ Species)
```

There is (a lot) more
---------------------

- More arguments to customize shapes (aesthetic mapping or otherwise).
- More geoms.
- Coordinate systems.
- Position adjustments.
- Statistical transformations.
- Themes.
- Packages that use ggplot2 (example: function `ggpairs()` from  `GGally`).