---
title: "Tools for Practice"
subtitle: "Statistical Computing with R"
author: "Waseem Medhat"
date: "`r Sys.Date()`"
logo: "content/r-logo.svg"
output:
  ioslides_presentation:
    wide: true
---

Help
----

- Functions are associated with help pages (documentation) containing
  information about arguments, expected inputs and outputs, resources on
  statistical method, etc.
- Simply, put `?` before something to access its help page (if it exists).

```{r eval=FALSE}
?which()
```

Vignettes
---------

- These are tutorials or demonstration documents that are associated with each
  packages.
- Use `vignette(package = "<package_name>")` to see available vignettes for any
  package
- Use `vignette("<vignette_name>")` to open the vignette.

```
library(dplyr)
vignette(package = "dplyr")
vignette("rowwise")
```

Missing data
------------

- Coded in R as `NA`.
- Can be in vectors (of any type) or data frames.

```{r}
c(9, 2, NA, 1, 8, NA, 5)
```

- Keep in mind the difference between `NA` and
    + `NULL` the absence of an object
    + `NaN` "not a number" might be due to erroneous calculations

```{r}
0 / 0
```


Importing SPSS data
-------------------

- R has a built-in package called `foreign` which allows importing files from
  other statistical packages: SPSS, SAS, etc.

```
library(foreign)
read.spss("<file_name.sav>", to.data.frame = TRUE)
```

Working with files
------------------

- R looks for files relative to the "current" directory, known as the _working
  directory_.

- You can use `getwd()` and `setwd()` to find or change the working directory,
  respectively.

- Instead of code, RStudio has options in the file browser to find and set the
  working directory.

Student's t-test
----------------

The `t.test()` function allows performing all the t-test family. The type of
test is determined by the arguments:

- `x`: dependent variable's observations of group 1 (numeric)
- `y`: dependent variable's observations of group 2 (numeric)
    + We won't provide an argument here in a 1-sample t-test.
    + both `x` and `y` can be replaced with a formula.
- `alternative`: test direction, either `"two.sided"`, `"less"`, or
  `"greater"`
- `mu`: population null value
- `paired`: whether or not the test is paired (logical)
- `var.equal`: whether or not to assume equal variance (logical)
- `conf.level`: confidence level as a decimal (< 1) not a percentage

Student's t-test
----------------

```{r}
# defaults
# var.equal = FALSE, alternative = "two.sided", mu = 0
# paired = FALSE,    conf.level = 0.95
t.test(x = iris$Sepal.Length[1:50],
       y = iris$Sepal.Length[51:100])
```

Chi-squared test
----------------

- Introducing a new function: `table()` which takes a data frame of two
  categorical variables and returns a contingency table.

```{r echo=FALSE, message=FALSE, include=FALSE}
knitr::knit("part_ii_basics.Rmd")
library(dplyr)
```

```{r}
disease_vs_older50 <- statisticians %>% 
  select(disease, older_than_50) %>% 
  table()
disease_vs_older50
```

Chi-squared test
----------------

Output from `table()` can be used as input to `chisq.test()`.

```{r}
chisq.test(disease_vs_older50)
```

Linear regression
-----------------

- Uses a concise formula syntax.
- No need to create dummy variables. R takes the (factor) predictor(s) and
  automatically creates the dumy variables.

```{r}
iris_model <- lm(Sepal.Length ~ Sepal.Width + Species, data = iris)
```

- The object `iris_model` contains the model, and it can be used with a wide
  range of functions that retrieve residuals, coefficients, diagnostics, etc.

- Example functions that work with an `lm` object: `summary()`, `residuals()`,
  `coef()`, `predict()`, `hatvalues()`, `cooks.distance()`

Linear regression output
------------------------

```{r eval=FALSE}
summary(iris_model)
```

```
Call:
lm(formula = Sepal.Length ~ Sepal.Width + Species, data = iris)

Residuals:
     Min       1Q   Median       3Q      Max 
-1.30711 -0.25713 -0.05325  0.19542  1.41253 

Coefficients:
                  Estimate Std. Error t value Pr(>|t|)    
(Intercept)         2.2514     0.3698   6.089 9.57e-09 ***
Sepal.Width         0.8036     0.1063   7.557 4.19e-12 ***
Speciesversicolor   1.4587     0.1121  13.012  < 2e-16 ***
Speciesvirginica    1.9468     0.1000  19.465  < 2e-16 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1
```

Linear regression output
------------------------

```
Residual standard error: 0.438 on 146 degrees of freedom
Multiple R-squared:  0.7259,	Adjusted R-squared:  0.7203 
F-statistic: 128.9 on 3 and 146 DF,  p-value: < 2.2e-16
```

Formula syntax
--------------

- Simple: `D ~ I`
- Multiple: `D ~ I1 + I2`
- Interaction term: `D ~ I1 + I2 + I3 * I4`
    + Using `*` adds `I3`, `I4` and the interaction between them.


R Markdown
----------

- A simple but elegant way to write documents.
- Can produce Word, PowerPoint, PDF documents, PDF slides, and HTML out of the
  box.
- Can be extended with packages to create books, websites, thesis dissertations,
  and dashboards.
- A document is written in some simple, readable syntax which is then "knitted"
  to the desired output.

R Markdown example
------------------

```
---
title: "Cross-Sectional Study of Smokers"
author: "Waseem Medhat"
date: 2020-12-31
---

# 1. Introduction

## 1.1. Background

Est quos *consectetur* ex. Inventore accusamus vero eum incidunt ad omnis eum
voluptates. Doloribus similique modi perspiciatis quia iusto quam dignissimos.

## 1.2 Current Research

Est quos consectetur ex. Inventore **accusamus** vero eum incidunt ad omnis eum
voluptates. Doloribus similique modi perspiciatis quia iusto quam dignissimos.
```