---
title: "Data Manipulation with dplyr"
subtitle: "Statistical Computing with R"
author: "Waseem Medhat"
date: "`r Sys.Date()`"
logo: "content/r-logo.svg"
output:
  ioslides_presentation:
    wide: true
    mathjax: local
    self_contained: false
---

What is `dplyr`?
----------------

<div class="red2">
"How can I work efficiently with data frames?"
</div>

- `dplyr` is most often the library of choice to work with data frames and
  compute variables, calculate summaries or grouped summaries, filter or sort
  some values, etc.  It offers a new mini-language with readable syntax.

    + `select()` to select columns.
    + `filter()` to filter rows meeting some (logical) criteria.
    + `mutate()` to compute new variables.
    + `summarize()` to compute summary statistics.
    + `group_by()` to stratify calculations.
    + `arrange()` to sort values.
    + The "pipe" operator `%>%`.

The pipe
--------

```{r}
# root mean squared error (RMSE)
errors <- c(1, 2, -1, 0, -3)
rmse <- sqrt(mean(errors ^ 2))
rmse
```

- This example used nested functions. Wouldn't it be easier to read if the syntax
  reflected the order of operations?

```
Square the errors --then-> calculate the mean --then-> take the square root
```

The pipe
--------

- The pipe operator `%>%` takes the output of its LHS and "pipes" it into the
  RHS function _as a first argument_.

```{r message=FALSE}
library(dplyr)
```

```{r}
(errors ^ 2) %>% mean() %>% sqrt()
```

The pipe
--------

- **Note**: the pipe works very well with `dplyr` because its functions take a
  data frame and return a data frame.

```
data frame with height and weight %>% 
  calculate BMI as new variable %>% 
  filter BMIs < 18 %>%
  calculate mean age
```

Select variables
----------------

```{r echo=FALSE, message=FALSE, include=FALSE}
knitr::knit("part_ii_basics.Rmd")
```

- Data frames can be piped into `select()` with the names of the desired
  arguments.

```{r}
names(statisticians)
```


```{r}
statisticians %>% select(name, age)
```

Remove variables
-----------------

- If you want to remove a variable instead of selecting it, put `-` before its
  name.

- This is useful when you have a large number of columns and you want all of
  them except a few.

```{r}
statisticians %>% select(-name)
```

Filter rows
-----------

- Data frames are piped into `filter()` with a comparison test (with or without
  boolean operators) as filter criteria.

```{r}
statisticians %>% filter(disease == "normal" & age < 40)
```

Compute new variables
---------------------

- `mutate()` allows you to derive new variables from existing ones.

```{r}
statisticians %>% mutate(age_z = (age - mean(age)) / sd(age))
```

Compute new variables
---------------------

- It can also replace existing variables if you use its name as the argument
  name.

```{r}
statisticians %>% mutate(name = toupper(name))
```

Compute summary statistics
--------------------------

- Aggregate functions (functions that take multiple numbers and return a single
  number, like `mean()`) can be called on variables with `summarize()`.

```{r}
statisticians %>% summarize(mean_age = mean(age))
```

- Multiple summaries can be computed simultaneously.

```{r}
statisticians %>% summarize(mean_age = mean(age),
                            sample_size = n()) # n() is special for summarize()
```

Grouped summaries
-----------------

- Data frames are piped into `group_by()` to make any subsequent calculations
  grouped/stratified by a certain grouping variable (or more).

```{r message=FALSE}
# calculate mean age in each disease group
statisticians %>% 
  group_by(disease) %>% 
  summarize(mean_age = mean(age))
```

Grouped summaries
-----------------

```{r message=FALSE}
# mean age by disease by age group
statisticians %>% 
  group_by(disease, older_than_50) %>% 
  summarize(mean_age = mean(age),
            sample_size = n())
```

Sort rows
---------

- `arrange()` sorts the rows of a data frame by one or more variables.

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(mean_wt = mean(wt)) %>% 
  arrange(mean_wt)
```

Sort rows
---------

- Wrapping a variable name in `desc()` sorts by descending order.

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarize(mean_wt = mean(wt)) %>% 
  arrange(desc(mean_wt))
```

A step back
===========

The power of `dplyr` 
---------------------

<div class="red2">
"What does `dplyr` add to my tool box that is not in, say, SPSS?"
</div>

- **Nondestructive**: The `dplyr` verbs always return new data frames. Notice that
  `statisticians` will always be intact unless something was explicitly assigned
  into it.
- **Flexibility**: operations can be combined and arranged in any way.
    + Warning: consider the order carefully as it might be a source of errors.

```
statisticians %>% 
  mutate(age_z = (age - mean(age)) / sd(age)) %>% 
  select(name, disease) %>% 
  filter(age_z < 50)  # error: object age_z not found
```


Errors
------

<div class="red2">
"Why is my code not working?"
</div>

- **Wrong syntax**:
    + forgot to close a parenthesis
    + extra commas
    + `=` instead of `==`
- **Wrong input type**:
    + `mean()` with a character
    + `filter()` with a vector
- Forgot to create a variable or load a package.
- Everything is correct to R but **wrong logic**.

